---
title: "Data Challenge 5"
author: "Xueqi Huang"
date: "December 04, 2021"
output:
  html_document:
    toc: yes
    toc_depth: 3
    toc_float: yes
---
Github: https://github.com/vickynugget/DSI_DataChallenge5.git

```{r setup, include=FALSE} 
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
```

```{r}
# load library
library(tidyverse)
library(janitor)
library(ggplot2)
library(GGally)
library(plotly)
```

## Loading/Cleaning Data and Exploratory Analysis
```{r}
# read data
nndb_flat <- read.csv('nndb_flat.csv')


nndb_flat <- nndb_flat %>%
  # filter the data to contain only specific food groups
  filter(FoodGroup == 'Vegetables and Vegetable Products' | 
           FoodGroup == 'Beef Products' | 
           FoodGroup == 'Sweets')

nndb_data <- nndb_flat %>%
  # select only the variables from Energy_kcal to Zinc_mg
  select(Energy_kcal:Zinc_mg)

# examine the correlation among the variables
ggcorr(nndb_data,        
       hjust = 1, 
       size = 2, 
       layout.exp = 2) # adjust label
```

Manganese and Vitamin A, Thiamin and Folate have the highest correlations and they are positive correlated. Zinc and protein, fat and energy, sugar and carb, Thiamin and Riboflavin also have high positive correlations. 


## Performing PCA
### 1. Perform PCA and scale the data
```{r}
# perform PCA and scale the data
pca_nndb <- prcomp(nndb_data, center = TRUE, scale. = TRUE)
```

### 2. Make a plot showing the cumulative proportion of the variation explained by each PC
```{r}
# Calculate the Cumulative proportion of the variation explained by each PC
var <- pca_nndb$sdev^2
cum <- cumsum(var/sum(var))

# Format into a data frame
var_explained_df <- data.frame(PC = paste0("PC",1:23),
                               var_explained = cum)

# format the order of PCs
var_explained_df$PC <- factor(var_explained_df$PC, levels = var_explained_df$PC)
var_explained_df %>%
  # initialize ggplot with PC on the x-axis and cumulative variation explained on the y-axis
  ggplot(aes(x = PC, y = var_explained, group = 1)) + 
  geom_point() + # add points to the plot
  geom_line() + # add lines to the plot
  labs(title = "Cumulative proportion of the variation explained", # add title
       y = 'Cumulated Variance Explained', # add y-axis label
       x = 'PC') + # add x-axis label
  theme(axis.text.x=element_text(angle = 40, hjust = 1)) # rotate the x labels
```

### 3. Make plots for the loadings on the first 3 PCs
```{r}
# get loadings for the first 3 PCs which explain about 60% of the variation in the data
pca_nndb_loadings <- as.data.frame(pca_nndb$rotation) %>% # get the loadings
  select(PC1, PC2, PC3) %>% # select first 3 PCs
  mutate(variable = rownames(pca_nndb$rotation)) %>% # add variable names
  pivot_longer(cols = c('PC1', 'PC2', 'PC3'), # transfer the data into longer format
               names_to = 'PC',
               values_to = 'loadings') 

# make 3 separate plots for the loadings for the first 3 PCs for all of the variables
for (i in c('PC1', 'PC2', 'PC3')){
  plot <- pca_nndb_loadings %>%
    filter(PC == i) %>% # filter out the loadings for specific PC
    ggplot(aes(x = reorder(variable, abs(loadings)), y = loadings)) + # initialize ggplot
    geom_bar(stat = 'identity') + # make a bar plot
    labs(title = paste("Loadings for", i), # add title
         y = 'Loadings', # add y-axis label
         x = 'Variable') + # add x-axis label
    theme(axis.text.x = element_text(angle=40, hjust=1)) # rotate the x labels
  # output the plot
  print(plot)
}
```

### 4. Make plots of the scores on the PCs
```{r}
# create a data frame of scores
pca_scores <- as.data.frame(pca_nndb$x) %>% # get the scores
  mutate(FoodGroup = nndb_flat$FoodGroup) # add the food group column

# PC1 versus PC2
pc1_pc2 <- ggplot(pca_scores, 
                  aes(x = PC1, y = PC2, col = FoodGroup)) + # initialize ggplot
  geom_point() + # make a scatter plot
  labs(title = 'PC1 vs PC2') # add a title
ggplotly(pc1_pc2) # interactive plot

# PC1 versus PC3
pc1_pc3 <- ggplot(pca_scores, 
                  aes(x = PC1, y = PC3, col = FoodGroup)) + # initialize ggplot
  geom_point() + # make a scatter plot
  labs(title = 'PC1 vs PC3') # add a title
ggplotly(pc1_pc3) # interactive plot

# PC2 versus PC3
pc2_pc3 <- ggplot(pca_scores, 
                  aes(x = PC2, y = PC3, col = FoodGroup)) + # initialize ggplot
  geom_point() + # make a scatter plot
  labs(title = 'PC2 vs PC3') # add a title
ggplotly(pc2_pc3) # interactive plot


```

## Identify Outlier and Performing PCA Again

The outlier is in the Vegetables and Vegetables Products group. 

### 1. Remove the outlier
```{r}
# find the index of the outlier, which has PC1 greater than 30
outlier_idx <- which(pca_scores[,'PC1'] > 30)

# remove the outlier
nndb_flat_new <- nndb_flat[-outlier_idx,]

nndb_data_new <- nndb_flat_new %>%
  # select only the variables from Energy_kcal to Zinc_mg
  select(Energy_kcal:Zinc_mg)
```

### 2. Perform PCA again
```{r}
# perform PCA and scale the data
pca_nndb <- prcomp(nndb_data_new, center = TRUE, scale. = TRUE)

# format proportion of the variation explained by each PC into a data frame
var <- pca_nndb$sdev^2
cum <- cumsum(var/sum(var))

# Format into a data frame
var_explained_df <- data.frame(PC = paste0("PC",1:23),
                               var_explained = cum)

# format the order of PCs
var_explained_df$PC <- factor(var_explained_df$PC, levels = var_explained_df$PC)

# Make a plot showing the cumulative proportion of the variation explained by each PC
var_explained_df %>%
  # initialize ggplot with PC on the x-axis and cumulative variation explained on the y-axis
  ggplot(aes(x = PC, y = var_explained, group = 1)) + 
  geom_point() + # add points to the plot
  geom_line() + # add lines to the plot
  labs(title = "Cumulative proportion of the variation explained", # add title
       y = 'Cumulated Variance Explained', # add y-axis label
       x = 'PC') + # add x-axis label
  theme(axis.text.x=element_text(angle = 40, hjust = 1)) # rotate the x labels


# get loadings for the first 3 PCs which explain about 60% of the variation in the data
pca_nndb_loadings <- as.data.frame(pca_nndb$rotation) %>% # get the loadings
  select(PC1, PC2, PC3) %>% # select first 3 PCs
  mutate(variable = rownames(pca_nndb$rotation)) %>% # add variable names
  pivot_longer(cols = c('PC1', 'PC2', 'PC3'), # transfer the data into longer format
               names_to = 'PC',
               values_to = 'loadings') 

# make 3 separate plots for the loadings for the first 3 PCs for all of the variables
for (i in c('PC1', 'PC2', 'PC3')){
  plot <- pca_nndb_loadings %>%
    filter(PC == i) %>% # filter out the loadings for specific PC
    ggplot(aes(x = reorder(variable, abs(loadings)), y = loadings)) + # initialize ggplot
    geom_bar(stat = 'identity') + # make a bar plot
    labs(title = paste("Loadings for", i), # add title
         y = 'Loadings', # add y-axis label
         x = 'Variable') + # add x-axis label
    theme(axis.text.x = element_text(angle = 40, hjust = 1)) # rotate the x labels
  # output the plot
  print(plot)
}


# make 3 plots of the scores on the PCs colored by food group
# create a data frame of scores
pca_scores <- as.data.frame(pca_nndb$x) %>% # get the scores
  mutate(FoodGroup = nndb_flat_new$FoodGroup) # add the food group column

# PC1 versus PC2
pc1_pc2 <- ggplot(pca_scores, 
                  aes(x = PC1, y = PC2, col = FoodGroup)) + # initialize ggplot
  geom_point() + # make a scatter plot
  labs(title = 'PC1 vs PC2') # add a title
ggplotly(pc1_pc2) # interactive plot

# PC1 versus PC3
pc1_pc3 <- ggplot(pca_scores, 
                  aes(x = PC1, y = PC3, col = FoodGroup)) + # initialize ggplot
  geom_point() + # make a scatter plot
  labs(title = 'PC1 vs PC3') # add a title
ggplotly(pc1_pc3) # interactive plot

# PC2 versus PC3
pc2_pc3 <- ggplot(pca_scores, 
                  aes(x = PC2, y = PC3, col = FoodGroup)) + # initialize ggplot
  geom_point() + # make a scatter plot 
  labs(title = 'PC2 vs PC3') # add a title
ggplotly(pc2_pc3) # interactive plot
```

Yes. Because we scale the data before performing PCA, so after remore the outliers the value after scaling will change. The loadings and score of PCA will change as a result. 

### Describe what you see 
From the plot of PC1 and PC2 loadings shows most of the loading were positive, while most loadings of PC3 were negative. So in the plot of scores, most scores of PC1 and PC2 lies in the first quadrant, in the plot of PC1 and PC3 most points lies below x-axis and left of the y-axis. In the plot of PC2 and PC3 most points lies below x-axis and right of y-axis. 
